Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-06-27 writing/DraftExposurePaper/DraftExposurePaper.Rmd
compare WU and NOAA best tracks data
Respond to GBA. Update count figures. Add to bib.

Current hurricane exposure assignments are often nonspecific [GBA-- I found this wording a bit unclear] and may lead to missclassification of exposure at the county-level [@Grabich2016]. Researchers have found heterogeneity in exposure assignments of counties, both within and between storms, when different metrics of exposure are used [@Grabich2016; @Grabich2015]. Such missclassification can have important consequences in assessments of the public health and economic impacts of tropical storms.

@@

Another study explored trends in direct hurricane fatalities for all landfalling US hurricanes between 1970 and 2007 in counties that were "exposed" to storms, based on a distance-based metric (whether the county was along the path of the storm [primary exposure], neighboring a county along the storm's path [secondary counties], or neighboring a secondary county [tertiary county]), determining wind- and rain-related hazards for all "exposed" counties and exploring associated with fatality counts [@Czajkowski2011].

@@

The NLDAS-2 precipitation data is originally provided hourly for a 1/8 degree grid [@Rui2014; @AlHamdan2014]. To generate county-level daily rainfall estimates, we first aggregated data at each grid point, after converting the timestamp of each observation to local time, to create a daily estimate at the grid point. We then averaged all grid points within a county's boundaries to generate a county-level average of daily precipitation [@AlHamdan2014]. This county-level precipitation data is publicly available through the US Centers for Disease Control's Wide-ranging Online Data for Epidemiological Research (WONDER) database [citation for WONDER; @AlHamdan2014].

Given its better consistancy with streamflow gage measurements compared to NLDAS precipitation data [GBA-- you had something in here about resolution, as well. Does it have a better resolution than the NLDAS data? I think we probably want to focus on comparing this source with NLDAS, and we don't need to put much about advantages that it has that are very similar to NLDAS], @Villarini2011 recommends Stage IV precipitation data for the analysis of rainfall distribution in landfalling tropical cyclones. However, the Stage IV dataset has only been archived since 2002 [@Lin2005] and cannot be used to analyze exposure to tropical storms for studies that include storms before 2002.

@@

```{r echo = FALSE}
data(closest_dist)
my_fips <- closest_dist$fips
my_fips <- unique(my_fips)
distance_exposure <- county_distance(counties = my_fips,
                                 start_year = 1988, end_year = 2011,
                                 dist_limit = 100)
distance_1000 <- distance_exposure %>%
        arrange(desc(storm_dist)) %>%
        group_by(fips) %>%
        summarise(value = n()) %>%
        rename(region = fips) %>%
        mutate(region = as.numeric(region), value = as.numeric(value)) %>%
        arrange(desc(value)) %>%
        slice(1:1000)
map_dist_1000 <- county_choropleth(distance_1000, 
                              title = "Hits by County\nfor Distance < 100 mm",
                              legend = "Hits per County", num_colors = 9,
                              state_zoom = eastern_states)
map_dist_1000 + scale_fill_brewer("Hits per\nCounty", palette = "YlGnBu", na.value = "white")
```

@@

```{r echo = FALSE, warning=FALSE}
rain_exposure <- county_rain(counties = my_fips,
                             start_year = 1988, end_year = 2011,
                             rain_limit = 75, dist_limit = 500,
                             days_include = c(-1, 0, 1))
rain_1000 <- rain_exposure %>%
        arrange(desc(tot_precip)) %>%
        group_by(fips) %>%
        summarise(value = n()) %>%
        rename(region = fips) %>%
        mutate(region = as.numeric(region), value = as.factor(value)) %>%
        arrange(desc(value)) %>%
        slice(1:1000)
eastern_states <- c("alabama", "arkansas", "connecticut", "delaware",
                    "district of columbia", "florida", "georgia", "illinois",
                    "indiana", "iowa", "kansas", "kentucky", "louisiana",
                    "maine", "maryland", "massachusetts", "michigan",
                    "mississippi", "missouri", "new hampshire", "new jersey",
                    "new york", "north carolina", "ohio", "oklahoma",
                    "pennsylvania", "rhode island", "south carolina",
                    "tennessee", "texas", "vermont", "virginia",
                    "west virginia", "wisconsin")
rain_1000$value <- as.numeric(rain_1000$value)
map_1000 <- county_choropleth(rain_1000,
                              title = "Hits for top 1000 Storm-County Pairs\nfor Rain > 75 mm",
                              legend = "Hits per County", num_colors = 9,
                              state_zoom = eastern_states)
map_1000 + scale_fill_brewer("Hits per\nCounty", palette = "YlGnBu", na.value = "white")
```

Map of total count of "exposures" by county, based on "exposed" = distance of 100 [or do we want a different distance limit for this? What seems reasonable?] kilometers or less:

```{r echo = FALSE}
distance_count <- distance_exposure %>%
        arrange(desc(storm_dist)) %>%
        group_by(fips) %>%
        summarise(value = n()) %>%
        rename(region = fips) %>%
        mutate(region = as.numeric(region), value = as.numeric(value)) %>%
        arrange(desc(value))
map_dist_count <- county_choropleth(distance_count, 
                              title = "Hits by County\nfor Distance < 100 km",
                              legend = "Hits per County", num_colors = 9,
                              state_zoom = eastern_states)
map_dist_count + scale_fill_brewer("Hits per\nCounty", palette = "YlGnBu", na.value = "white")
```

@@

While there are often several different "best tracks" datasets, from different sources [?-- weather services?] for other ocean basins [@Knapp2010], the "best tracks" data from NOAA's National Hurricane Center [@Jarvinen1988], which are the basis of the extended hurricane tracks data we use here, are the undisputed primary source of tropical storm track data for the Atlantic basin [@Knapp2010]. One study [@Currie2013] did, however, use hurricane track data from the Weather Underground Hurricane Archive [GBA-- is Weather Underground just using the NOAA best tracks? That's my guess. In that case, this source would be the same].

>>>>

The most often used hurricane exposure criteria at the county-level include FEMA disaster declarations and area within the storm track, which can be somewhat nonspecific and potentially lead to misclassification of exposure at the county-level [@Grabich2016]. FEMA declaration is largely subjective, requiring state governors and/or tribal chiefs to submit a proposal for government aid on the basis that local resources are overwhelmed [@FEMA]. As a result, using FEMA disaster declarations as exposure classification to hurricanes restricts exposure to 1) those states whose resources are overwhelmed and 2) those states who apply for FEMA aid. As a result, other severely impacted counties and states that may not fit FEMA criteria are excluded and may therefore be incorrectly identified as unexposed, preventing researchers from assessing the full impact of a given hurricane. Researchers have found heterogeneity in exposure assignments of counties, both within and between storms, when different metrics of exposure are used [@Grabich2016; @Grabich2015]. Such missclassification can have important consequences in assessments of the public health and economic impacts of tropical storms.

@@

Another study explored trends in direct hurricane fatalities for all landfalling US hurricanes between 1970 and 2007 in counties that were "exposed" to storms, based on a distance-based metric (whether the county was along the path of the storm [primary exposure], neighboring a county along the storm's path [secondary counties], or neighboring a secondary county [tertiary county]), determining wind- and rain-related hazards for all "exposed" counties and exploring associations with fatality counts [@Czajkowski2011].

@@

The NLDAS-2 precipitation data is originally provided hourly for a 1/8 degree grid [@Rui2014; @AlHamdan2014]. To generate county-level daily rainfall estimates, we first aggregated data at each grid point, after converting the timestamp of each observation to local time, to create a daily estimate at the grid point. We then averaged all grid points within a county's boundaries to generate a county-level average of daily precipitation [@AlHamdan2014]. This county-level precipitation data is publicly available through the US Centers for Disease Control's Wide-ranging Online Data for Epidemiological Research (WONDER) database [@WONDER; @AlHamdan2014].

Given its better consistancy with streamflow gage measurements compared to NLDAS precipitation data [GBA-- you had something in here about resolution, as well. Does it have a better resolution than the NLDAS data? I think we probably want to focus on comparing this source with NLDAS, and we don't need to put much about advantages that it has that are very similar to NLDAS; JMF-- Page 5 in Villarini 2011 disucsses how the spacial resolution is much finer for Stage IV than NLDAS and TMPA, which they argue causes a smoothing of maximum rainfall data for the Appalachians], @Villarini2011 recommends Stage IV precipitation data for the analysis of rainfall distribution in landfalling tropical cyclones. However, the Stage IV dataset has only been archived since 2002 [@Lin2005] and cannot be used to analyze exposure to tropical storms for studies that include storms before 2002.

@@

```{r echo = FALSE}
data(closest_dist)
my_fips <- closest_dist$fips
my_fips <- unique(my_fips)
distance_exposure <- county_distance(counties = my_fips,
                                 start_year = 1988, end_year = 2011,
                                 dist_limit = 100)
distance_1000 <- distance_exposure %>%
        arrange(desc(storm_dist)) %>%
        group_by(fips) %>%
        summarise(value = n()) %>%
        rename(region = fips) %>%
        mutate(region = as.numeric(region), value = as.numeric(value)) %>%
        arrange(desc(value)) %>%
        slice(1:1000)
eastern_states <- c("alabama", "arkansas", "connecticut", "delaware",
                    "district of columbia", "florida", "georgia", "illinois",
                    "indiana", "iowa", "kansas", "kentucky", "louisiana",
                    "maine", "maryland", "massachusetts", "michigan",
                    "mississippi", "missouri", "new hampshire", "new jersey",
                    "new york", "north carolina", "ohio", "oklahoma",
                    "pennsylvania", "rhode island", "south carolina",
                    "tennessee", "texas", "vermont", "virginia",
                    "west virginia", "wisconsin")
map_dist_1000 <- county_choropleth(distance_1000, 
                              title = "Hits by County\nfor Distance < 100 km",
                              legend = "Hits per County", num_colors = 9,
                              state_zoom = eastern_states)
map_dist_1000 + scale_fill_brewer("Hits per\nCounty", palette = "YlGn", na.value = "white")
```

@@

```{r echo = FALSE, warning=FALSE}
rain_exposure <- county_rain(counties = my_fips,
                             start_year = 1988, end_year = 2011,
                             rain_limit = 75, dist_limit = 500,
                             days_include = c(-1, 0, 1))
rain_1000 <- rain_exposure %>%
        arrange(desc(tot_precip)) %>%
        group_by(fips) %>%
        summarise(value = n()) %>%
        rename(region = fips) %>%
        mutate(region = as.numeric(region), value = as.factor(value)) %>%
        arrange(desc(value)) %>%
        slice(1:1000)
rain_1000$value <- as.numeric(rain_1000$value)
map_1000 <- county_choropleth(rain_1000,
                              title = "Hits for top 1000 Storm-County Pairs\nfor Rain > 75 mm",
                              legend = "Hits per County", num_colors = 9,
                              state_zoom = eastern_states)
map_1000 + scale_fill_brewer("Hits per\nCounty", palette = "YlGnBu", na.value = "white")
```

Map of total count of "exposures" by county, based on "exposed" = distance of 100 [or do we want a different distance limit for this? What seems reasonable?; JMF-- Based on what I have read so far, I think this seems reasonable, unless we want to push it out to 120 km to loosely reflect the Czajkowski study which seems to have found a great number of fatalities in tertiary counties] kilometers or less:

```{r echo = FALSE}
distance_count <- distance_exposure %>%
        arrange(desc(storm_dist)) %>%
        group_by(fips) %>%
        summarise(value = n()) %>%
        rename(region = fips) %>%
        mutate(region = as.numeric(region), value = as.numeric(value)) %>%
        arrange(desc(value))
map_dist_count <- county_choropleth(distance_count, 
                              title = "Hits by County\nfor Distance < 100 km",
                              legend = "Hits per County", num_colors = 9,
                              state_zoom = eastern_states)
map_dist_count + scale_fill_brewer("Hits per\nCounty", palette = "YlGn", na.value = "white")
```

@@

While there are often several different "best tracks" datasets, from different sources [?-- weather services?] for other ocean basins [@Knapp2010], the "best tracks" data from NOAA's National Hurricane Center [@Jarvinen1988], which are the basis of the extended hurricane tracks data we use here, are the undisputed primary source of tropical storm track data for the Atlantic basin [@Knapp2010]. One study [@Currie2013] did, however, use hurricane track data from the Weather Underground Hurricane Archive [GBA-- is Weather Underground just using the NOAA best tracks? That's my guess. In that case, this source would be the same. JMF-- they have an archive from 1851-2016; I think it is the same source. I'll keep digging though].
